**The Heart Lies: The Role of Interoception and Theory of Mind in Deception**

### 2022-11-01 paper/manuscript.Rmd
manu

The fabrication and delivery of false information is a sophisticated, and cognitively taxing activity [@abe2007deceiving], resulting in physiological arousal [@eskritt2021did] which may require effortful self-regulation to conceal [@krokoszinski2016emotion]. While much of deception research has focused on topics such as theory of mind, other mechanisms underlying the regulation of physiological responses, such as increased cardiac arousal, respiration and sweating, which are often relied on during lie detection have been under researched.

@@

Interoception refers to a cluster of skills involved in monitoring, interpreting, integrating, and regulating one's visceral states [@chen2021emerging; @weiss2014interaction], and is closely tied to emotional and cognitive processes [@barrett2015interoceptive; @petzschner2021computational]. @garfinkel2015knowing consolidated the facets of interoception into three distinct processes: *interoceptive accuracy*, as an objective measure of one's ability to monitor viscera; *interoceptive sensibility*, as the subjective measure of accuracy; and *interoceptive awareness*, describing the metacognition underlying the perception of interoceptive accuracy. While interoceptive sensibility may be an important component, the corrospondence between the confidence and accuracy, or interoceptive awareness appears to be associated with the downregulation of emotional responses and their associated physiological symptoms [@fustos2013embodiment; @critchley2017interoception; @critchley2018influence].

@@

Interoceptive processes are not only a key component of forming your own emotions, but may also facilitate being in tune with the emotional states of others [@konvalinka2011synchronized; @garfinkel2016discrepancies]. This corresponds to situations involving deception, since much of the existing literature examining the link between interoceptive abilities has focused on how it may implicitly flag that one is being lied to through an elevated cardiac response [@gunderson2021body], and skin temperature [@van2014registered]. While research indicates that attending to ones interoceptive cues improves success in unmasking detection [@ten2019different], there is limited information how interoceptive skills may function the other way around, in the case of the liar.

As identified by @murphy2022propensity, the propensity to use interoceptive information, and in what manner remains an underdeveloped area in research, and may illuminate the mechanisms underlying physiological and emotional control exhibited by successful liars [@krokoszinski2016emotion; @wielgopolan2021can].

@@

This research attempts to isolate the effects of theory of mind and interoception to examine how participants may rely on these abilities to evade lie detection. Of particular interest is how liars might utilize their interoceptive ability to self regulate in towards a more successful deception, and how this ability may compare or interact with other mechanisms at play such as theory of mind. We hypothesize that participants with greater theory of mind ability will show higher lie confidence, shorter reaction time, and lower physiological arousal in the interrogation condition, while participants in the polygraph condition will demonstrate lower lie confidence, lower reaction time and higher physiological arousal.

@@

Participants underwent 40 trails of a directed lying task, where they were asked questions such as, "What secondary school did you go to?". Answers were provided verbally and participant pressed a button when they were finished, at which point they would be connected for 10 seconds to the condition-unique stimulus. There were two conditions of 20 trails each, beginning with the *interrogation* *condition*, and followed by the second, *polygraph condition*. The purpose of these conditions was to draw out the interoceptive or theory of mind abilities by providing feedback cues that would cause them to focus their attention on either the visual feedback cues from other people, or their own cardiac response.

>>>>

In our current post-truth era, deception has become increasingly prevalent in our everyday lives [@lewandowsky2017beyond]. Although the act of deceiving is complex and cognitively taxing [@abe2007deceiving], often requiring effortful self-regulation to conceal [@krokoszinski2016emotion], research shows that people tend to under-estimate their ability to successfully deceive others [@elaad2003effects]. Moreover, meta-analytic findings suggest that the ability of the liar plays a more significant role than the lie detector in deceptive interactions [@verigin2019lie; @bond2008individual; @levine2011sender]. However, relative to the abundant literature regarding deception detection [@viji2022history; @sternglanz2019review; @masip2017deception], surprisingly little attention has been conferred to investigating the inter-individual differences in people's ability to successfully lie.

<!-- TOM  -->

As deception requires the liar to intentionally manipulate the beliefs of others [@burgoon1994interpersonal; @sip2012if], most research on lying capability has been focused on higher order cognitive processes, particularly in theory of mind (TOM). TOM refers to the ability to infer that others have mental states, such as beliefs, emotions and intentions, distinct from ourselves [@lee2021lying; @wellman2001meta; @baron1997mindblindness]. Indeed, the capacity to tell lies, as well as the complexity of lies told, have previously been found to be associated with higher TOM abilities [@talwar2007lying; @evans2011verbal; @talwar2017role]. Furthermore, the deceptive capabilities of children has often been used as a measure of their TOM abilities. However, given that the development of TOM abilities has often been viewed as the building blocks of other higher order cognitive processes [@ketelaars2010dynamics; @isaksson2021s], studies investigating the links between TOM and lying ability have largely been conducted with children and individuals with neurocognitive deficits [@beaudoin2020systematic; @bora2017meta; @roheger2022progression]. As such, little is known with regards to the role TOM plays in the lying ability of neurotypical adults.

@@

Alternatively, the variability observed in interoception, broadly defined as one's sensitivity to their own internal signals and bodily states [@murphy2019classifying; @chen2021emerging; @weiss2014interaction], has increasingly been tied to subjective perceptual experiences [@connell2018interoception; @seth2012interoceptive], as well as individual differences in executive functions, emotional processing and decision-making [@barrett2015interoceptive; @petzschner2021computational; @murphy2019classifying]. Following @garfinkel2015knowing's three-dimensional framework, interoception is comprised of distinct facets, namely - *interoceptive accuracy*, the objective ability to monitor internal bodily signals; *interoceptive sensibility*, the subjective confidence in one's interoceptive accuracy; and *interoceptive awareness*, the metacognitive awareness of one's interoceptive accuracy (i.e., accuracy-confidence correspondence).

<!-- While interoceptive sensibility may be an important component, the correspondence between the confidence and accuracy, or interoceptive awareness appears to be associated with the downregulation of emotional responses and their associated physiological symptoms [@fustos2013embodiment; @critchley2017interoception; @critchley2018influence]. -->

@@

Given the influence of interoceptive abilities on higher order cognition, particularly with regards to emotional processing  [@barrett2015interoceptive; @petzschner2021computational; @vabba2022interoceptive], it is possible that interoception bears an effect on people's capacity to tell lies. Although there has been little research investigating the links between interoception and deceptive ability *per se*, some studies have found heightened self-focus towards internal bodily signals to result in an increased demonstration of apathetic, immoral behaviour, such as cheating [@ditto2006visceral; @williams2016valjean; @lenggenhager2013listening]. In line with these findings, @vabba2022interoceptive reports individuals with lower interoception told significantly less egoistic lies when the social reputational stakes were high (relative to low), whereas individuals with higher interoception did not exhibit a significant difference in the number of lies told.

In contrast, however, the somatic marker hypothesis posits higher interoception to be linked to a propensity towards making rational decisions [@damasio1996somatic]. Indeed, previous research has shown a negative correlation between interoceptive awareness and the likelihood to make risky decisions [@dunn2010listening; @furman2013interoceptive]. Taking these mixed findings in perspective with the relative lack of deception research associated with the multidimensional construct of interoception, more studies are warranted to investigate the relationship between the interoceptive and deceptive ability.

<!-- Interoceptive processes are not only a key component of forming your own emotions, but may also facilitate being in tune with the emotional states of others [@konvalinka2011synchronized; @garfinkel2016discrepancies]. This corresponds to situations involving deception, since much of the existing literature examining the link between interoceptive abilities has focused on how it may implicitly flag that one is being lied to through an elevated cardiac response [@gunderson2021body], and skin temperature [@van2014registered]. While research indicates that attending to ones interoceptive cues improves success in unmasking detection [@ten2019different], there is limited information how interoceptive skills may function the other way around, in the case of the liar. -->

<!-- As identified by @murphy2022propensity, the propensity to use interoceptive information, and in what manner remains an underdeveloped area in research, and may illuminate the mechanisms underlying physiological and emotional control exhibited by successful liars [@krokoszinski2016emotion; @wielgopolan2021can]. -->

@@

This study aims at exploring the relationship between TOM and interoception with deceptive ability. Despite the criticism surrounding the somatic marker hypothesis [@dunn2006somatic; @maia2004reexamination], interoception has been shown to be associated with goal-directed behaviours and positively influence rational decision-making in ambiguous situations [@poppa2018somatic; @naqvi2014insula; @avery2017brain]. As such, we expect individuals higher in interoception to exhibit better lying ability, as indicated by higher lie confidence, lower physiological arousal and shorter response time in the polygraph condition. In the interrogation condition, we expect people with higher TOM abilities to exhibit be better at lying, as indicated by higher low confidence, lower physiological arousal and shorter response time.

<!-- This research attempts to isolate the effects of theory of mind and interoception to examine how participants may rely on these abilities to evade lie detection. Of particular interest is how liars might utilize their interoceptive ability to self regulate in towards a more successful deception, and how this ability may compare or interact with other mechanisms at play such as theory of mind. We hypothesize that participants with greater theory of mind ability will show higher lie confidence, shorter reaction time, and lower physiological arousal in the interrogation condition, while participants in the polygraph condition will demonstrate lower lie confidence, lower reaction time and higher physiological arousal. -->

@@

Participants underwent 40 trails of a directed lying task, where they were asked questions such as, "What secondary school did you go to?". Answers were provided verbally and participant pressed a button when they were finished, at which point they would be connected for 10 seconds to the condition-unique stimulus. There were two conditions of 20 trails each, beginning with the *interrogation condition*, and followed by the second, *polygraph condition*. The purpose of these conditions was to draw out the interoceptive or theory of mind abilities by providing feedback cues that would cause them to focus their attention on either the visual feedback cues from other people, or their own cardiac response.
